The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth

### 2021-09-20 paper/paper-jsdse/paper-jsdse.tex
fixed begin{document} not found issue

In this section, we outline the steps taken to the download the raw data from the database provided at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data} (The U.S. Bureau of Labor Statistics, n.d.). Specifically, we only extract a subset of the variables in the database as shown in \ref{box:data}.

\hypertarget{getting-and-tidying-the-data}{%
\subsection{Getting and Tidying the Data}\label{getting-and-tidying-the-data}}

@@

\faDatabase National Longitudinal Survey of Youth 1979

@@

@@

{\centering \includegraphics[width=468px]{figures/flow-chart-1}

@@

\hypertarget{eda}{%
\section{Exploratory Data Analysis}\label{eda}}

This section will provide an example of how data sets could be used to do exploratory data analysis as in textbook data. We are interested in examining how the respondents experience changes in their wages based on their demographic characteristics. Thus, we use \texttt{wages} data instead of the refreshed \texttt{wages\_hs\_do} as it only has male respondents.

The statistics we use in measuring the change in wages experience is the slope of the wages of each respondent.
Using \texttt{key\_slope} function in \texttt{brolgar} (Tierney, Cook, and Prvan 2020), we perform linear regression with the log of mean hourly wage as the response variable and year as the predictor. Since we do modeling, it is important to look at the number of observations from each respondent. Figure \ref{fig:slope-nobs-plot} shows that the number of observations for each individual varies. Moreover, each number of observations seems to have an impact on the slope. Extreme values in slope are observed in a few observations, while the range of slope is more stable with more observations. Accordingly, we trim the data to only include respondents with at least 10 observations.

{\centering \includegraphics[width=468px]{figures/slope-nobs-plot-1}

\caption{A) Distribution of observations. The numbers of rounds varies among respondents. Most of respondents participated in 28 rounds of survey. B) Distribution of wages slope by number of observations. The slopes' range are more stable with more observations.}\label{fig:slope-nobs-plot}
\end{figure}

We then create some visualizations of five respondents with the highest slope and five respondents with the lowest slope, henceforth referred to as top-5 slopes and bottom-5 slopes, respectively. We plot them based on their gender and education. Figure \ref{fig:slope-gender} shows top-5 slopes are dominated by males, while females dominate bottom-5 slopes. However, as this figure only shows a small portion of the data, we further inspect the gender wage gap.

{\centering \includegraphics[width=468px]{figures/slope-gender-1}

\caption{Wages slope over the years based on gender. Most of respondents who experienced high wages slope are males. In contrast, respondents who experienced lower wages slope are mostly females.}\label{fig:slope-gender}
\end{figure}

According to OECD (n.d.), the gender pay gap is the difference between the median wage of males and females relative to the median wage of males. We find that, as shown in \ref{fig:gap-gender}, over the years, females tend to earn fewer than males. For example, in 2018, females earned about 78 cents for every \$1 earned by males. The widest gender gaps were observed in the 1990s. However, as more males participated in the survey than females from year to year, these gaps could be affected by imbalance number of observations.

{\centering \includegraphics[width=0.9\linewidth]{figures/gap-gender-1}

\caption{A) Number of observations by gender and by year. There are more males participated in survey than females. The number of observations decreased over the years. B) Gender wage gap or the median wages of females relatives to median wages of males. Females tend to earn fewer than males. For example, in 2018 females earn about 78 cents for every \$1 earned by males. Note that this is a crude measure as gap is calculated not based on other variable, such as experience and job class.}\label{fig:gap-gender}
\end{figure}

According to their highest education completed, both top-5 and bottom-5 slopes are dominated by people who completed high school. This is understandable since the respondents are mostly completed high school.

{\centering \includegraphics[width=468px]{figures/slope-educ-1}

\caption{Wages slope over the years based on highest education completed. People who are in top-5 and bottom-5 wages are dominated by people who completed high school.}\label{fig:slope-educ}
\end{figure}

Further, we perform a linear regression to see the effect of education on wages experience.

@@

\leavevmode\hypertarget{ref-OECD}{}%
OECD. n.d. {``Gender Wage Gap.''} https://data.oecd.org/earnwage/gender-wage-gap.htm.

>>>>

In this section, we outline the steps taken to the download the raw data from the database provided at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data} (The U.S. Bureau of Labor Statistics, n.d.). Specifically, we only extract a subset of the variables in the database as described in Section\textasciitilde{}\ref{sec:getdata}.

\hypertarget{getdata}{%
\subsection{Getting and Tidying the Data}\label{getdata}}

@@

\faDatabase~National Longitudinal Survey of Youth 1979

@@

\hypertarget{recap}{%
\subsection{Recap}\label{recap}}

@@

{\centering \includegraphics[width=0.93\linewidth]{/Users/etan0038/Dropbox/repo/summer-wages-refresh/paper/figures/flowchart}

@@

\hypertarget{exploratory-data-analysis}{%
\section{Exploratory Data Analysis}\label{exploratory-data-analysis}}

@@

### 2021-09-20 paper/paper.tex
fixed begin{document} not found issue

\usepackage{tcolorbox}
\newtcolorbox{tbox}{width=\textwidth,boxsep=5pt}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\hypertarget{getting-and-tidying-the-data}{%
\subsection{Getting and Tidying the Data}\label{getting-and-tidying-the-data}}

@@

Education, Training and Achievement Scores

\begin{itemize}
\tightlist
\item
  Education -\textgreater{} Summary measures -\textgreater{} All schools -\textgreater{} By year -\textgreater{} Highest grade completed

  \begin{itemize}
  \tightlist
  \item
    Downloaded all of variables in Highest grade completed.
  \end{itemize}
\end{itemize}

Employment

\begin{itemize}
\tightlist
\item
  Summary measures -\textgreater{} By job -\textgreater{} Hours worked and Hourly wages

  \begin{itemize}
  \tightlist
  \item
    Downloaded all of the variables in Hours worked.
  \item
    Downloaded all of the variables in Hourly wages
    Both hours worked and hourly wages are recorded by the job, up to five jobs for each id/subject.
  \end{itemize}
\end{itemize}

Household, Geography and Contextual Variables

\begin{itemize}
\tightlist
\item
  Context -\textgreater{} Summary measures -\textgreater{} Basic demographics

  \begin{itemize}
  \tightlist
  \item
    Downloaded year and month of birth, race, and sex variables.
    There are two versions of the year and month of birth, i.e., 1979 and 1981 data. We downloaded these two versions.
  \end{itemize}
\end{itemize}

>>>>

\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

In this section, we outline the steps taken to the download the raw data from the database provided at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data} (The U.S. Bureau of Labor Statistics, n.d.). Specifically, we only extract a subset of the variables in the database as described in Section\textasciitilde{}\ref{sec:getdata}.

\hypertarget{getdata}{%
\subsection{Getting and Tidying the Data}\label{getdata}}

@@

\begin{tcolorbox}[label="box:data", title=Getting the data]

\faDatabase~National Longitudinal Survey of Youth 1979

\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year $\triangleright$ Highest grade completed
\begin{itemize}
\item[\faCloudDownload] Downloaded all of variables in Highest grade completed.
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job $\triangleright$ Hours worked and Hourly wages
\begin{itemize}
\item[\faCloudDownload] Downloaded all of the variables in Hours worked.
\item[\faCloudDownload] Downloaded all of the variables in Hourly wages
  Both hours worked and hourly wages are recorded by the job, up to five jobs for each id/subject. 
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography and Contextual Variables
\begin{itemize}
\item[$\triangleright$] Context $\triangleright$ Summary measures $\triangleright$ Basic demographics 
\begin{itemize}
\item[\faCloudDownload] Downloaded year and month of birth, race, and sex variables. There are two versions of the year and month of birth, i.e., 1979 and 1981 data. We downloaded these two versions.
\end{itemize}
\end{itemize}
\end{itemize}

\end{tcolorbox}
